---
title: "Australian Bushfires"
author: "Tom Maccor"
format: html
editor: visual
knitr: 
  opts_chunk:
    message: false
    warning: false
---

# Statistical analysis, modelling and initial forecasting

This paper will aim at answering the question:

Is there a statistically significant increase in the incidence (nr. fires/year) and size of fires in New South Wales Australia, since bushfire data started being recorded?

## [1.- Data Collection]{.underline}

An internet search for data sources was performed. The datasets that showed most promise were the ones kept by the [NSW Department of Climate Change, Energy, the Environment and Water](https://datasets.seed.nsw.gov.au/dataset/fire-history-wildfires-and-prescribed-burns-1e8b6) (click for the URL).

The datasets are in Shapefile format (.shp), which is a simple, nontopological format for storing the geometric location and attribute information of geographic features. Geographic features in a shapefile can be represented by points, lines, or polygons (areas), which is our case for the size of area burnt each year.

Now, the dataset obtained had to be converted into a format that can be read by the **R** programming language.

This was done via [QGIS](https://qgis.org/en/site/index.html), an open source Geospatial software.

Once a **.csv** file was obtained, it was read into R:

```{r}
#| echo: false

libraries <- c('tidyverse', 'lubridate', 'readxl', 'MARSS', "forecast", 'stats', 'gridExtra', 'feasts', 'stats', 'fable', 'plotly', 'tseries', 'urca', 'reshape2')

lapply(libraries, require , character.only = TRUE)

```

```{r}
NSW_fire_history <- read_csv("NSW_fire_history_2024.csv")

glimpse(NSW_fire_history)
```

We can see that there is data on bushfires for the period 1899-2024:

```{r}
datetest <- NSW_fire_history$StartDate %>% na.omit ()
range(datetest)
```

but..... there are a lot of missing records:

```{r}
sum(is.na(NSW_fire_history$StartDate))
```

15,850 missing Start Dates out of a total of 35.776 records. That is too much.

A histogram shows that, not surpringingly, there is more are more data recorded in recent years versus the 1st half of the 20th century:

```{r}
#| echo: false

histo <- ggplot(NSW_fire_history, mapping = aes(StartDate)) + geom_histogram()
histo
```

Which are the statistical tests that can be used to determine if there is statistically significant increase in the data trend?

## 

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}

## Analize data from 1953 onwards  ---> too little data prior to this year
## 2. So will filter by "Label" variable
## First filter rows for "wildfire"  (removing prescribed burns)
## then by year   

```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

ou to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
